Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-26 paper.tex
Fixed issue #8.

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=false]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract software components
  and interactions in 6 architectures that we use to classify 18
  systems. We compare these architectures using metrics measuring
  integration effort, robustness, extensibility, scalability and other
  specific features. Results show the pros and cons of the different
  architectures: tight integration and sub-tasking require the least
  amount of integration effort and are also the most robust, most
  architectures perform the same in terms of extensibility, the pool
  model is the most scalable and meta-workflows are only available in
  nested workflows and workflow import. These results provide insights
  for software architects on how to integrate science gateways and
  workflow engines.
\end{abstract}

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
integration effort, robustness, extensibility, scalability and other
specific features of the architectures. We use these metrics to
evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine architects.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration effort, robustness of workflow
execution, extensibility, scalability and other specific
features. Criteria break down to specific metrics where \emph{lower
  value indicates better performance}. For each criterion, a total
score is computed by summing up the individual metrics. We ensure that
the different metrics in a criterion measure similar entities so that
they can be summed. The criteria and metrics are explained hereafter.

@@

\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\

@@

\emph{Integration effort} measures the development required to build
the architecture. It is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for integration effort, which measures the total number of
software pieces to develop (pieces include interactions and
components).

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be more prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interactions and components that need to be modified when a new
element is added. Modification of a component is required when its
code needs to be updated or recompiled (science gateway or workflow
service), or when a new piece of software has to be installed
(infrastructure only). Modification of an interaction is deemed
necessary when the parameters involved in this interaction are
modified.  Extensibility breaks down in 4 metrics depending on the
type of element that has to be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain). 
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

\paragraph{Integration effort} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

@@

\paragraph{Integration effort} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

@@

\paragraph{Integration effort} The sub-task architecture requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Integration effort} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

@@

\paragraph{Integration effort} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (appears twice: once for each infrastructure),
\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
(twice: once for each workflow engine and infrastructure) and
\texttt{d} (twice: once for each workflow engine).
\answerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}

@@

\paragraph{Integration effort} Workflow import with service invocation
requires the same components as for service invocation
(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
for workflow import (\texttt{I$_1$}=7).

@@

Tight integration and sub-tasking require the least amount of
integration effort, followed by service integration, workflow import
(with service invocation) and pool. Nested workflows (with service
invocation) requires more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and sub-tasking in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics, it reflects the global complexity of the
architectures.

@@

  Integration effort (normalized total) &

@@

Moreover, all the interactions involved in the architectures were
treated equally, as if they all required similar amounts of
development effort, which is realistically not the case. For example,
interaction \texttt{g} (workflow language conversion) clearly requires
more effort than interaction \texttt{d} (service invocation), and may
create much more reliability issues as well. However, quantifying the
robustness and amount of development associated with each interaction
and software component cannot realistically be done without entering
into the details of a particular system.

@@

% Summary of contributions
We reviewed architectures used to integrate workflow engines in
science gateways, we described them in a system-independent framework
suitable for comparison, we illustrated them on real systems, we
developed quantitative metrics for evaluation, we evaluated these
metrics for all the architectures, and we discussed the pros and cons
of all the architectures based on these metrics.  \answerednote{Rafael}{One
  thing that is missing in the conclusion or the limitations section
  is to state that this work does not intend to quantify the amount of
  work required to develop science gateways using a specific
  architecture, but the goal is to highlight what are the main
  components and interactions for each architecture, and the main
  advantages and drawbacks. This should also be stated in the
  introduction.}{Tristan}{This is in contradiction with the metric on
  'integration effort'. What do you think?}

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=false]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\closedanswerednote}[6]{\pdfmargincomment[color=gray,author=#1-#3-#5,open=true]{#1: #2 \textCR \textCR #3: #4 \textCR \textCR #5: #6}}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. Based on our experience with
  the development and operations of the CBRAIN and VIP science
  gateways, and our past involvement in the SHIWA project, we abstract
  software components and interactions in 6 architectures that we use
  to classify 18 systems. We compare these architectures using metrics
  measuring global complexity, robustness, extensibility, scalability
  and other specific features. Results show the pros and cons of the
  different architectures: tight integration and sub-tasking are the
  simplest architectures and should therefore be the easiest to
  implement and the most robust ones, most architectures perform the
  same in terms of extensibility, the pool model is the most scalable
  and meta-workflows are only available in nested workflows and
  workflow import. These results provide insights for software
  architects on how to integrate science gateways and workflow
  engines.
\end{abstract}

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
global complexity, robustness, extensibility, scalability and other
specific features of the architectures. We use these metrics to
evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine architects.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: global complexity, execution complexity,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\cellcolor[HTML]{EEEEEE}\textbf{Global complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\

@@

\emph{Global complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for global complexity. Global complexity measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity since complex architectures tend to be more prone to
failure. More precisely, robustness is determined as the number of
software components and interactions that are specific to workflow
execution:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are the ones represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the total score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain). 
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

\paragraph{Global complexity} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

@@

\paragraph{Global complexity} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

@@

\paragraph{Global complexity} The sub-task architecture requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Global complexity} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

@@

\paragraph{Global complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (appears twice: once for each infrastructure),
\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
(twice: once for each workflow engine and infrastructure) and
\texttt{d} (twice: once for each workflow engine).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.}

@@

\paragraph{Global complexity} Workflow import with service invocation
requires the same components as for service invocation
(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
for workflow import (\texttt{I$_1$}=7).

@@

Tight integration and sub-tasking are globally the simplest
architectures, followed by service integration, workflow import (with
service invocation) and pool. Nested workflows (with service
invocation) requires more integration than the other
architectures. Robustness roughly leads to the same ordering of
architectures, with tight integration and sub-tasking in the top
group, service integration and workflow import (with service
invocation) close behind, pool in a third group, and nested workflows
(with service invocation) at the end. This ordering is consistent
across metrics, it reflects the global complexity of the
architectures.

@@

  Global complexity (normalized total) &

@@

Moreover, all the interactions involved in the architectures were
treated equally while some of them are obviously more complex
than others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may create much more reliability issues as
well. However, quantifying the robustness and amount of development
associated with each interaction and software component cannot
realistically be done without entering into the details of a
particular system.

@@

% Summary of contributions
We reviewed architectures used to integrate workflow engines in
science gateways, we described them in a system-independent framework
suitable for comparison, we illustrated them on real systems, we
developed quantitative metrics for evaluation, we evaluated these
metrics for all the architectures, and we discussed the pros and cons
of all the architectures based on these metrics.
\closedanswerednote{Rafael}{One thing that is missing in the
  conclusion or the limitations section is to state that this work
  does not intend to quantify the amount of work required to develop
  science gateways using a specific architecture, but the goal is to
  highlight what are the main components and interactions for each
  architecture, and the main advantages and drawbacks. This should
  also be stated in the introduction.}{Tristan}{This is in
  contradiction with the metric on 'integration effort'. What do you
  think?}{Tristan}{Fixed, as discussed in github \#8.}
